---
title: "Hands-on Exercise 06: Geographical Segmentation with Spatially Constrained Clustering"
author: "Meng Xinyu"
date: "`r Sys.Date()`"
date-modified: "last-modified"
format: html
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **12  Geographical Segmentation with Spatially Constrained Clustering Techniques**

## **12.1 The data**

### **12.1.1 Installing and loading R packages**

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```

## **12.2 Data Import and Prepatation**

### **12.2.1 Importing geospatial data into R environment**

```{r}
shan_sf <- st_read(dsn = "data/geospatial",
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

```{r}
shan_sf
```

```{r}
glimpse(shan_sf)
```

### **12.2.2 Importing aspatial data into R environment**

```{r}
ict <- read_csv("data/aspatial/Shan-ICT.csv")
```

```{r}
summary(ict)
```

### **12.2.3 Derive new variables using dplyr package**

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

```{r}
summary(ict_derived)
```

## **12.3 Exploratory Data Analysis (EDA)**

### **12.3.1 EDA using statistical graphics**

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

What can you observed from the distributions reveal in the histogram and boxplot？

The histograms reveal that the six ICT penetration rate variables differ substantially in both their distribution shape and their scale. TV_PR is the most symmetric and roughly bell-shaped, while RADIO_PR is wide and moderately right-skewed and represents the most widely adopted device. In contrast, LLPHONE_PR, COMPUTER_PR and INTERNET_PR are strongly right-skewed, with most townships clustered near low values and only a few reaching high penetration, indicating that landline phones, computers and home internet remain uncommon across most of Shan State. MPHONE_PR shows a spread, multi-peaked distribution, reflecting large variation in mobile phone adoption between townships. Two observations are important for the clustering that follows. First, the variables operate on very different numeric scales — RADIO_PR and TV_PR reach several hundred while COMPUTER_PR and INTERNET_PR only reach the tens — so the data must be standardised before clustering, otherwise the large-scale variables would dominate the distance calculations. Second, the pronounced right skew in the computer and internet variables highlights a clear digital divide within the region, which is exactly the kind of spatial heterogeneity the clustering analysis aims to uncover.

```{r}
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

### **12.3.2 EDA using choropleth map**

```{r}
#12.3.2.1 Joining geospatial data with aspatial data
shan_sf <- left_join(shan_sf, ict_derived, by = c("TS_PCODE" = "TS_PCODE"))

dir.create("data/rds", recursive = TRUE, showWarnings = FALSE)
write_rds(shan_sf, "data/rds/shan_sf.rds")
```

```{r}
shan_sf <- read_rds("data/rds/shan_sf.rds")
```

```{r}
#12.3.2.2 Preparing a choropleth map
qtm(shan_sf, "RADIO_PR")
```

```{r}
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks", 
          title = "Total households") + 
  tm_borders(alpha = 0.5) 

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

```{r}
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style="jenks") +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(outer.margins=0, asp=0)
```

Can you identify the differences?

Yes. The two choropleth maps show clearly different spatial patterns. In the left map, the townships with the highest total number of households (dark blue) are concentrated in the central part of the region, reflecting where the population is largest. In the right map, the townships with the highest radio penetration rate are located differently — mainly in the northwest and parts of the southwest — and several of the high-household townships in the centre actually have only moderate or low radio penetration. This mismatch confirms why the penetration rates were derived earlier: the raw counts are driven mainly by population size, so a township with many households will naturally own more radios, TVs and so on. By converting the counts into penetration rates per thousand households, we remove the effect of population size and capture the actual level of ICT adoption. The fact that the two maps do not align demonstrates that a township's size is not the same as its level of ICT development, which is exactly why the clustering analysis should be based on the penetration rates rather than the absolute counts.

## **12.4 Correlation Analysis**

```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

## **12.5 Hierarchy Cluster Analysis**

### **12.5.1 Extracting clustering variables**

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)

dir.create("data/rds", recursive = TRUE, showWarnings = FALSE)
write_rds(shan_ict, "data/rds/shan_ict.rds")
```

```{r}
shan_ict <- read_rds("data/rds/shan_ict.rds")
```

### **12.5.2 Min-Max standardisation**

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

### **12.5.3 Z-score standardisation**

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

### **12.5.4 Visualising the standardised clustering variables**

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

What statistical conclusion can you draw from the histograms above?

The three histograms share an identical shape — same modes, same right-skew, same relative bar heights. Standardisation (both Min-Max and Z-score) only rescales the x-axis: it makes variables measured on different units comparable so that no single variable dominates the Euclidean distance, but it does not change the underlying distribution and does not make the data more normal. Since RADIO_PR remains skewed after standardisation, Z-score (which assumes approximate normality) is less appropriate here than Min-Max normalisation.

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

### **12.5.5 Computing proximity matrix**

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
```

```{r}
proxmat
```

### **12.5.6 Computing hierarchical clustering**

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

```{r}
plot(hclust_ward, cex = 0.6)
```

### **12.5.7 Selecting the optimal clustering algorithm**

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

### **12.5.8 Determining Optimal Clusters**

```{r}
#12.5.8.1 Gap Statistic Method
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

```{r}
fviz_gap_stat(gap_stat)
```

### **12.5.9 Interpreting the dendrograms**

```{r}
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

### **12.5.10 Visually-driven hierarchical clustering analysis**

```{r}
#12.5.10.1 Transforming the data frame into a matrix
shan_ict_mat <- data.matrix(shan_ict)
```

```{r}
#12.5.10.2 Plotting interactive cluster heatmap using heatmaply()
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

### **12.5.11 Mapping the clusters formed**

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

```{r}
qtm(shan_sf_cluster, "CLUSTER")
```

## **12.6 Spatially Constrained Clustering: SKATER approach**

### **12.6.1 Converting into SpatialPolygonsDataFrame**

```{r}
shan_sp <- as_Spatial(shan_sf)
```

### **12.6.2 Computing Neighbour List**

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

```{r}
coords <- st_coordinates(
  st_centroid(st_geometry(shan_sf)))
```

```{r}
plot(st_geometry(shan_sf), 
     border=grey(.5))
plot(shan.nb,
     coords, 
     col="blue", 
     add=TRUE)
```

### **12.6.3 Computing minimum spanning tree**

```{r}
#12.6.3.1 Calculating edge costs
lcosts <- nbcosts(shan.nb, shan_ict)
```

```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

### **12.6.4 Computing minimum spanning tree**

```{r}
shan.mst <- mstree(shan.w)
```

```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

```{r}
head(shan.mst)
```

```{r}
plot(st_geometry(shan_sf), 
                 border=gray(.5))
plot.mst(shan.mst, 
         coords, 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

### **12.6.5 Computing spatially constrained clusters using SKATER method**

```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2], 
                 data = shan_ict, 
                 method = "euclidean", 
                 ncuts = 5)
```

```{r}
str(clust6)
```

```{r}
ccs6 <- clust6$groups
ccs6
```

```{r}
table(ccs6)
```

```{r}
plot(st_geometry(shan_sf), 
     border=gray(.5))
plot(clust6, 
     coords, 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink"),
     cex.circles=0.005, 
     add=TRUE)
```

### **12.6.6 Visualising the clusters in choropleth map**

```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

```{r}
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(alpha = 0.5) 

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```

## **12.7 Spatially Constrained Clustering: ClustGeo Method**

### **12.7.1 Ward-like hierarchical clustering: ClustGeo**

```{r}
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, 
            k = 6, 
            border = 2:5)
```

```{r}
#12.7.1.1 Mapping the clusters formed
groups <- as.factor(cutree(nongeo_cluster, k=6))
```

```{r}
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

### **12.7.2 Spatially Constrained Hierarchical Clustering**

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

```{r}
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K=6, graph = TRUE)
```

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
```

```{r}
groups <- as.factor(cutree(clustG, k=6))
```

```{r}
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_Gcluster, "CLUSTER")
```

## **12.8 Visual Interpretation of Clusters**

### **12.8.1 Visualising individual clustering variable**

```{r}
ggplot(data = shan_sf_ngeo_cluster,
       aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

### **12.8.2 Multivariate Visualisation**

```{r}
ggparcoord(data = shan_sf_ngeo_cluster, 
           columns = c(17:21), 
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) + 
  theme(axis.text.x = element_text(angle = 30))
```

```{r}
shan_sf_ngeo_cluster %>% 
  st_set_geometry(NULL) %>%
  group_by(CLUSTER) %>%
  summarise(mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```

# What I Learned

This exercise walked through segmenting the townships of Shan State by their ICT penetration rates, and it clarified why spatial context matters when clustering geographically referenced data. I first used correlation analysis to drop redundant variables — COMPUTER_PR and INTERNET_PR were highly correlated, so keeping both would let the same signal dominate the distance matrix. Comparing raw, min-max and z-score histograms made a subtle point concrete: standardisation only rescales a variable's axis so that no single variable dominates the Euclidean distance, but it does not reshape the distribution or make skewed data normal, which is why min-max is safer than z-score here. Conventional hierarchical clustering (Ward's method on a proximity matrix) groups townships purely on attribute similarity, and mapping those clusters showed them scattered across the state rather than forming coherent regions. Spatially constrained approaches fix this: SKATER prunes a minimum spanning tree of contiguous neighbours so every cluster is geographically connected, while ClustGeo blends attribute distance and spatial distance through an alpha parameter, trading a little within-cluster homogeneity for spatial compactness. The overall lesson is that the "best" clustering statistically is not always operationally useful — for regionalisation, contiguity is a real constraint, and choosing between SKATER and ClustGeo is a deliberate trade-off rather than a purely technical choice.
