---
title: "Forest-fire Point Patterns in Kab. Pulang Pisau"
subtitle: "Take-home Exercise 1 · Executive Summary · ISSS626"
author: "Meng Xinyu"
date: "2026-09-23"
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    incremental: false
    footer: "ISSS626 Geospatial Analytics · Take-home Ex1"
execute:
  echo: false
  eval: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
# Rebuild the objects needed for the plots below (same pipeline as the report).
pacman::p_load(sf, tmap, spatstat, raster, tidyverse)

# boundary / study window
study_area <- st_read("data/raw", layer = "gadm41_IDN_2", quiet = TRUE) %>%
  filter(NAME_2 == "Pulang Pisau") %>%
  st_transform(crs = 32749)

# fire detections (two files: standard + near-real-time)
fire_archive <- read_csv("data/raw/fire_archive_J1V-C2_811410.csv",
                         col_types = cols(.default = col_character()))
fire_nrt     <- read_csv("data/raw/fire_nrt_J1V-C2_811410.csv",
                         col_types = cols(.default = col_character()))
fire_raw <- bind_rows(archive = fire_archive, nrt = fire_nrt, .id = "source") %>%
  type_convert()

fire_clean <- fire_raw %>%
  distinct() %>%
  filter(!is.na(latitude), !is.na(longitude),
         between(latitude, -90, 90), between(longitude, -180, 180)) %>%
  filter(confidence %in% c("n", "h")) %>%
  mutate(acq_date = as.Date(acq_date)) %>%
  filter(acq_date >= as.Date("2026-01-01"), acq_date <= as.Date("2026-09-23")) %>%
  mutate(month = lubridate::floor_date(acq_date, "month"))

fire_sf <- fire_clean %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 32749) %>%
  st_intersection(study_area)

fire_dedup <- fire_sf %>%
  mutate(x = st_coordinates(.)[, 1], y = st_coordinates(.)[, 2]) %>%
  distinct(x, y, acq_date, .keep_all = TRUE)

fire_owin <- as.owin(study_area)
fire_ppp  <- as.ppp(st_geometry(fire_dedup)); Window(fire_ppp) <- fire_owin
fire_ppp  <- rjitter(fire_ppp, retry = TRUE, nsim = 1, drop = TRUE)
fire_ppp.km <- rescale.ppp(fire_ppp, 1000, "km")

# first-order intensity
kde_ppl <- density(fire_ppp.km, sigma = bw.ppl, edge = TRUE, kernel = "gaussian")

# subsample for the second-order envelope (fewer sims here to keep rendering quick)
set.seed(1234)
fire_sub <- fire_ppp.km[sample(fire_ppp.km$n, 3000)]
L_env <- envelope(fire_sub, Lest, nsim = 9, correction = "Ripley",
                  global = FALSE, verbose = FALSE)
```

## Contents

1.  Research question & study area
2.  Data & study design
3.  Analytical approach
4.  First-order: intensity
5.  Second-order: interaction
6.  Robustness: inhomogeneous L
7.  Value-added: timing
8.  Key findings
9.  Limitations
10. Public-safety implications

::: notes
Cover + contents do not count toward the 10-slide limit.
:::

# Research question & study area

## Where do fires concentrate — and do they cluster?

-   Peatland fires in Central Kalimantan cause haze, health and economic harm.
-   **Q1** Where does fire-detection intensity concentrate, and when?
-   **Q2** Are detections significantly clustered, beyond just being dense?
-   Study window: **Kabupaten Pulang Pisau**, a peat-dominated regency; 2026.

```{r}
#| fig-height: 3.2
tm_shape(study_area) + tm_polygons(alpha = 0.3) +
  tm_layout(frame = FALSE)
```

# Data & study design

## Data & definitions

-   **Detections** — NASA FIRMS VIIRS 375 m (NOAA-20), 1 Jan – 23 Sep 2026.
-   **Boundary** — GADM ADM2 (regency), projected to UTM 49S.
-   **Point event** = one detection; a detection is **not** a distinct fire, so repeats are de-duplicated to one point per pixel per day.
-   Cleaning: 370,354 → 351,822 national records; **16,189** inside the regency.

# Analytical approach

## Methods

| Question    | Method                     | Compared against   |
|-------------|----------------------------|--------------------|
| Intensity   | Kernel density (KDE)       | —                  |
| Interaction | G / F / K / L functions    | homogeneous CSR    |
| Robustness  | Inhomogeneous L            | inhomogeneous null |
| Timing      | Detections over the season | —                  |

# First-order: intensity

## Detections concentrate in the south

```{r}
#| fig-height: 3.6
kde_rast <- raster(kde_ppl); crs(kde_rast) <- "EPSG:32749"
tm_shape(kde_rast) +
  tm_raster(palette = "YlOrRd", title = "Intensity") +
  tm_shape(study_area) + tm_borders(col = "grey30") +
  tm_layout(frame = FALSE, legend.outside = TRUE)
```

**Intensity is highly uneven** — a clear hotspot in the central-south peatland, stable across bandwidths.

# Second-order: interaction

## Strong clustering vs a random baseline

```{r}
#| fig-height: 3.6
plot(L_env, . - r ~ r, main = "", legend = FALSE)
abline(h = 0, lty = 2)
```

The observed L-curve sits **far above** the CSR envelope up to \~20 km → **significant clustering** at all scales.

# Robustness: inhomogeneous L

## Most clustering is the hotspot, not interaction

-   Under a homogeneous null, detections look strongly clustered.
-   But once **intensity is allowed to vary** (inhomogeneous L), the curve falls back near the baseline.
-   **Interpretation:** the clustering mostly reflects **first-order heterogeneity** — fires are concentrated because the conditions for fire are concentrated — not extra fire-to-fire interaction.

# Value-added: timing

## Activity peaks as the dry season advances

```{r}
#| fig-height: 3.4
fire_dedup %>%
  st_drop_geometry() %>%
  count(month) %>%
  ggplot(aes(month, n)) +
  geom_col(fill = "firebrick") +
  labs(x = NULL, y = "Detections") +
  theme_minimal(base_size = 14)
```

Detections rise through the year toward the dry-season peak; the hotspot stays central-south but **intensifies over time**.

# Key findings

## What we learned

-   **First-order:** intensity highly uneven, concentrated in the central-south peatland.
-   **Second-order:** significant clustering under CSR — but the **inhomogeneous L shows it is mostly first-order heterogeneity**.
-   **Timing:** activity concentrates in the dry-season months.
-   First- and second-order evidence agree: a few peat-dominated areas drive the pattern.

# Limitations

## Read with care

-   Detections ≠ fires; cloud cover and overpass timing leave observation gaps.
-   Concentration of detections ≠ risk to people, without exposure data.
-   CSR / KDE assumptions hold only partly; results read against more than one null.
-   Window stops in September — the later dry-season peak is not captured.

# Public-safety implications

## From evidence to action

-   **Where:** prioritise monitoring, patrols and response resources in the central-south peatland.
-   **When:** raise fire-danger readiness ahead of the dry-season peak.
-   **Longer term:** peatland water management (maintaining levels, canal blocking) in the hotspot.
-   Directly supported: **spatial concentration + timing**. Needs more data: causation, who is exposed.
